\documentclass[english, inz, shortabstract]{iithesis}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

\author{Mikołaj Depta}
\advisor{dr Andrzej Łukaszewski}
\transcriptnum {328690}
\advisorgen    {dr. Andrzeja Łukaszewskiego}
\englishtitle{Analysis of Type-Driven approach to systems programming:\fmlinebreak Implementation of OpenGL library for Rust}

\polishtitle{Analiza programowania systemowego z wykorzystaniem systemu typów:\fmlinebreak Implementacja biblioteki do OpenGL dla języka Rust}

\englishabstract{For the past few years in the software development industry there has been a growing interest in strongly typed languages. It manifests itself in emergence of brand-new technologies in which strong type systems were one of the core founding principles or in changes introduced to existing languages. The most common examples of modern languages with powerful type systems are TypeScript as an alternative to JavaScript in the world of web development or Rust in domain of systems programming in place of C and C++. More mature languages also had their type systems revised for example in C\# 8 explicit type nullability annotations were introduced, or even dynamically typed Python has seen major improvements to its type annotation system.

This study - an implementation of the OpenGL graphics API wrapper library for Rust - will attempt to demonstrate how Rust’s type system can be utilized to improve low-level software safety and maintainability as well as how it affects API design and codebase structure.
}

\polishabstract{For the past few years in the software development industry there has been a growing interest in strongly typed languages. It manifests itself in emergence of brand-new technologies in which strong type systems were one of the core founding principles or in changes introduced to existing languages. The most common examples of modern languages with powerful type systems are TypeScript as an alternative to JavaScript in the world of web development or Rust in domain of systems programming in place of C and C++. More mature languages also had their type systems revised for example in C\# 8 explicit type nullability annotations were introduced, or even dynamically typed Python has seen major improvements to its type annotation system.

This study - an implementation of the OpenGL graphics API wrapper library for Rust - will attempt to demonstrate how Rust’s type system can be utilized to improve low-level software safety and maintainability as well as how it affects API design and codebase structure.
}

\begin{document}

\include{chapter-1/main}
\include{chapter-2/main}
\include{chapter-3/main}

\bibliographystyle{plain}
\bibliography{bibliography}

\end{document}
